The effects of context inference on motor learning: savings, de-adaptation and spontaneous recovery

### 2021-08-31 adaptation/article/article.tex

\subsection{Extended explanatory power}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context, including infinite. It is in these cases where the effects of context inference are most noticeable. Laboratory experiments with this property are common. While many experiments exist that give probabilistic cues [EGs], or cues that only reveal part of the context [EGs], evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

@@

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is gated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1). It is important to note that they did not calculate a learning of exactly zero in these near-maximal-error trials, which suggests that participants are doing \textit{post hoc} learning: after observing the outcome of their actions, they identify that they had inferred the context incorrectly, and can then adapt the movements for the correct context.

A subtler source of information can be found in long pauses between blocks of adaptation trials. \cite{Ethier_Spontaneous_2008} performed an experiment where the adaptation blocks were divided into sub-blocks of 60 trials each, with a long 30s pause between sub-blocks. The authors observed that at the beginning of each sub-block, participants showed an average dip in their learned adaptation, suggesting that they were partially returning to baseline (O) after each pause. This can be explained by context inference, as a long pause could prompt participants to infer that a switch had occured. Because this was inferred without any information that points at the identity of the new context, prior beliefs take over. This means that participants rely on their belief of the underlying probability of observing any of the known contexts, which is dominated by the previously observed context A, but now includes a component of the baseline O, as it is the most common in their lives.

Error-clamp (E) trials present another insight. During E trials, error is artificially kept at zero. This takes different forms, depending on the type of experiment. If error is kept at zero, one could assume that participants would continue doing what they were doing before, as there is no reason (no observed error) to inferr a change in context. However, this is almost never the case \cite[e.g.][]{Smith_Interacting_2006,Ethier_Spontaneous_2008,Forano_Timescales_2020,Vaswani_Decay_2013,Scheidt_Persistence_2000}. Instead, participants slowly reduce their adaptation, often displaying spontaneous recovery \cite[e.g.][]{Smith_Interacting_2006}, i.e. an initial response in the direction of one of the learned adaptations. Our model provides a principled account of this behavior: while the observed error (i.e. the cursor on the screen) is kept at zero, participants still receive proprioceptive feedback which tells them that something has changed. The natural variability in a participant's behavior will lead her to expect observing errors, which clashes with the observed zero error. This prompts the participant to re-evaluate their inferred context, which can partially activate a previously-learned adaptation, as we showed in \fref{fig:vaswani-2013}. This can be seen in \cite{Criscimagna-Hemminger_Consolidation_2008}, where the authors showed that introducing long periods before the E block begins lowers the initial force that participants exerted on the mechanical arm during the E block; longer periods of time make context inference revert to the prior expectation that a new baseline block begins, because participants are free to move their arm about during the pause \todo{check this}.

>>>>

\subsection{Extended explanatory power}
Context is often directly observable, either via environmental cues or via feedback observed after actions. In many behavioral experiments, for example, the context of the current block of trials is easily identifiable by the color of a cue light \citep[e.g.][]{Ethier_Spontaneous_2008}, the posture of the hand \citep[e.g.][]{Gandolfo_Motor_1996}, the target of the movement \citep[e.g.][]{Lee_Dual_2009} or by the workspace in which the task is being performed \citep[e.g.][]{Shadmehr_Adaptive_1994}. The context can also be selected by the agent itself. For example, a person can decide the level of care devoted to moving an egg or a lead ball around, which determines the way in which the person would move their hands. In these cases, the context is clearly determined and known to the agent and switches between different adaptations are immediate.

In other cases, however, the context might not be directly observable, and context inference takes the form of an evidence-accumulating process that can take any amount of time to be sure of the context. It is in these cases where the effects of context inference are most noticeable. While many experiments exist that give probabilistic contextual information \cite[e.g.][]{Scholz_uncontrolled_1999,Behrens_Learning_2007,Nassar_Dissociable_2019}, evidence accumulation is not limited to these explicit cases. Indeed, as we argued in the Results section, many experiments inadvertently include partial contextual information used by participants.

@@

The same rationale explains the results from \cite{Davidson_Scaling_2004}: motor learning, which in our model is gated by context inference, is minimal for errors close to 2 and -2 (see their figure 2E). This is because an error of 2 or -2 signals that the participant incorrectly identified the context (as adaptation has a magnitude of 1).

A subtler source of information can be found in long pauses between blocks of adaptation trials. \cite{Ethier_Spontaneous_2008} performed an experiment where the adaptation blocks were divided into sub-blocks of 60 trials each, with a long 30s pause between sub-blocks. The authors observed that at the beginning of each sub-block, participants showed an average dip in their learned adaptation, suggesting that they were partially returning to baseline (O) after each pause. This can be explained by context inference, as a long pause could prompt participants to infer that a switch had occured. Because this was inferred without any information that points at the identity of the new context, prior beliefs take over. This means that participants rely on their belief of the underlying probability of observing any of the known contexts, which is dominated by the previously observed context A, but now includes a component of the baseline O, as it is the most common one in everyday life.

Error-clamp (E) trials present another insight. During E trials, error is artificially kept at zero. This takes different forms, depending on the type of experiment. If error is kept at zero, one could assume that participants would continue doing what they were doing before, as there is no reason (no observed error) to inferr a change in context. However, this is almost never the case \cite[e.g.][]{Smith_Interacting_2006,Ethier_Spontaneous_2008,Forano_Timescales_2020,Vaswani_Decay_2013,Scheidt_Persistence_2000,Pekny_Protection_2011}. Instead, participants slowly reduce their adaptation, often displaying spontaneous recovery \cite[e.g.][]{Smith_Interacting_2006}, i.e. an initial response in the direction of one of the learned adaptations. Our model provides a principled account of this behavior: while the observed error (i.e. the cursor on the screen) is kept at zero, participants still receive proprioceptive feedback which tells them that something has changed. The natural variability in a participant's behavior will lead her to expect observing errors, which clashes with the observed zero error. This prompts the participant to re-evaluate their inferred context, which can partially activate a previously-learned adaptation, as we showed in \fref{fig:vaswani-2013}. \cite{Pekny_Protection_2011} found similar results, demonstrating that the duration of the previously-observed adaptation block also affects behavior in the E block. Additionally, \cite{Criscimagna-Hemminger_Consolidation_2008} showed that introducing long periods before the E block begins lowers the initial force that participants exerted on the mechanical arm during the E block; longer periods of time make context inference revert to the prior expectation that a new baseline block begins, because participants are free to move their arm about during the pause.
